Well-quasi-orderings on word languages \\ main edb3a6bc24be4a12fe4f383c75384a6dff9d1a3c \\ 2025-10-02 17:15:30 +0200

### 2025-10-13 src/infixes.tex
fix: correct typos and error noticed by Lia

\begin{lemma}
    \label{pumping-periods:lem}
    Let $L \subseteq \Sigma^*$ be a language
    that is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Let $k \in \Nat$, $u_1, \cdots, u_{k+1} \in \Sigma^*$,
    and $v_1, \cdots, v_{k} \in \Sigma^+$
    be such that
    $w[\vec{n}] \defined (\prod_{i = 1}^k u_i v_i^{n_i}) u_{k+1}$
    belongs to $L$
    for arbitrarily large values of $\vec{n} \in \Nat^k$.
    Then, 
    there exists $x,y \in \Sigma^+$ of size 
    at most $\max \setof{\card{v_i}}{1 \leq i \leq k}$
    such that
    for all \todo{proof does not actually show for all}
    $\vec{n} \in \Nat^{k}$
    one of the following holds:
    \begin{enumerate}
        \item $w[\vec{n}] \in u_1 \InfPeriodChain{x}$,
        \item $w[\vec{n}] \in \InfPeriodChain{x} u_{k+1}$,
        \item $w[\vec{n}] \in \InfPeriodChain{x} u_i \InfPeriodChain{y}$
            for some $1 \leq i \leq k + 1$.

@@

    Let us construct a sequence of words $\seqof{w_i}[i \in \Nat]$, where $w_i
    \defined w[\vec{n_i}]$ for some well-chosen indices $\vec{n_i} \in \Nat^k$. The goal
    being that 
    if $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$,
    then it can intersect at most \emph{two} iterated words,
    with an intersection that is long enough to successfully apply
    \cref{periodic-infixes:lem}.
    In order to achieve this,
    let us first define $s$ as the maximal size of a word $v_i$
    ($1 \leq i \leq k$) and $u_j$ ($1 \leq j \leq k+1$).
    Then,
    we consider $\vec{n_0} \in \Nat^k$ such that $\vec{n_0}$ has all 
    its components greater than $\factorial{s}$ and such that
    $w[\vec{n_0}]$ belongs to $L$. 
    Then, we inductively define 
    $\vec{n_{i+1}}$  as the smallest vector of numbers greater than $\vec{n_i}$,
    such that $w[\vec{n_{i+1}}]$ belongs to $L$, 
    and with $\vec{n_i}$ having all components greater than
    $2\card{w[\vec{n_i}]}$.

@@

@@

\begin{corollary}
    \label{bounded-wqo-dwclosed:cor}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the
    \kl{infix relation} if and only if $\dwset[\infleq]{L}$ is.
\end{corollary}
\begin{proof}
    Because $L \subseteq \dwset[\infleq]{L}$, the right-to-left implication
    is trivial.
    For the left-to-right implication, let us assume that $L$ is a
    \kl{well-quasi-ordered} language for the \kl{infix relation}.
    Then $L$ is included in a finite union 
    of products of \kl{chains}: \todo{reference}
    \begin{equation*}
        L \subseteq \bigcup_{i = 1}^n S_i \cdot P_i \quad .
    \end{equation*}
    Remark that the \kl{downwards closure} of a product of two \kl{chains}
    is a finite union of products of two chains. \todo{reference/proof}.
    As a consequence, we conclude that $\dwset[\infleq]{L}$ is itself included
    in a finite union of products of \kl{chains}.
    Note that this also proves that $\dwset[\infleq]{L}$ is a \kl{bounded language},
    hence that it is \kl{well-quasi-ordered} by the \kl{infix relation} 
    via
    \cref{bounded-language:thm}.
\end{proof}

One may think that all \kl{downwards closed} languages for the \kl{infix
relation} that are \kl{well-quasi-ordered} are \kl(language){bounded}. Note
that this is what happens in the case of the \kl{subword embedding}, where any
\kl{downwards closed} language for this relation is characterized by finitely
many excluded \kl{subwords}, hence provides a \kl{bounded language} \todo{wrong? $\{a,b\}^*$ is downward closed but not bounded}. However, this
is not the case for the \kl{infix relation}, as we will now illustrate with the
following two examples.

\begin{example}
    \label{dwclosed-wqo-not-finite-excl:ex}
    Let $L \defined a^* b^* \cup b^* a^*$ \todo{However, this language is bounded, so this seems to be a non sequitur}. This language is \kl{downwards
    closed} for the \kl{infix relation}, is \kl{well-quasi-ordered} for the
    \kl{infix relation}, but is characterized by an \emph{infinite} number 
    of excluded infixes, respectively of the form $ab^ka$ and $ba^kb$ where $k \geq 1$.
\end{example}

@@

    %Let us prove that these are upper bounds for the \kl{ordinal invariants} of
    $\LMorse$. 
    The bound of the \kl{ordinal height} holds for any language $L$,
    since a descending chain for $(\Sigma^*, \infleq)$ is isomorphic to one for $(\Nat, \leq)$ \todo{please check this is a good argument}.
    For the \kl{maximal order type}, we remark that
    the \kl{uniform recurrence} of $\ThueMorse$ means that
    the maximal length of a \kl{bad sequence} is determined by its first element,
    hence that it is at most $\omegaOrd$.
    Finally, because the \kl{ordinal width} is at most the \kl{maximal order type} \todo{add reference},
    we conclude that the \kl{ordinal width} is also at most $\omegaOrd$.

@@

\AP A subset $I \subseteq X$ is \intro(subset){directed} if, for every $x,y \in
I$, there exists $z \in I$ such that $x \leq z$ and $y \leq z$. Given a
\kl{well-quasi-order} $(X, \leq)$, one can always decompose $X$ into a finite
union of \intro{order ideals}, that is, non-empty sets $I \subseteq X$ that are
\kl{downwards closed} and \kl(subset){directed} for the relation $\leq$. In our
case, a \kl{well-quasi-ordered} \kl{order ideal} for the \kl{infix relation} is
the set of finite \kl{infixes} of a finite \todo{$\Sigma^\Nat$ is an infinite word}
or bi-infinte word $w \in
\Sigma^\Nat \cup \Sigma^\Rel $ (\cref{bi-infinite:lem}). Therefore, we will
first study languages that are described as the set of finite \kl{infixes} of
some (bi-)infinite word (\cref{ultimately-uniformly-recurrent:lem}). To that
end, let us introduce the notation $\intro*\infset{w}$ for the set of finite
\kl{infixes} of a (possibly infinite or bi-infinite) word $w$.

>>>>

\begin{lemma}
    \label{pumping-periods:lem}
    Let $L \subseteq \Sigma^*$ be a language
    that is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Let $k \in \Nat$, $u_1, \cdots, u_{k+1} \in \Sigma^*$,
    and $v_1, \cdots, v_{k} \in \Sigma^+$
    be such that
    $w[\vec{n}] \defined (\prod_{i = 1}^k u_i v_i^{n_i}) u_{k+1}$
    belongs to $L$
    for arbitrarily large values of $\vec{n} \in \Nat^k$.
    Then, 
    there exists $x,y \in \Sigma^+$ of size 
    at most $\min \setof{\card{v_i}}{1 \leq i \leq k}$
    such that
    for all \todo{proof does not actually show for all}
    $\vec{n} \in \Nat^{k}$
    one of the following holds:
    \begin{enumerate}
        \item $w[\vec{n}] \in u_1 \InfPeriodChain{x}$,
        \item $w[\vec{n}] \in \InfPeriodChain{x} u_{k+1}$,
        \item $w[\vec{n}] \in \InfPeriodChain{x} u_i \InfPeriodChain{y}$
            for some $1 \leq i \leq k + 1$.

@@

    Let us construct a sequence of words $\seqof{w_i}[i \in \Nat]$, where $w_i
    \defined w[\vec{n_i}]$ for some well-chosen indices $\vec{n_i} \in \Nat^k$. The goal
    being that 
    if $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$,
    then it can intersect at most \emph{two} iterated words,
    with an intersection that is long enough to successfully apply
    \cref{periodic-infixes:lem}.
    In order to achieve this,
    let us first define $s$ as the minimal size of a word $v_i$
    ($1 \leq i \leq k$) and $u_j$ ($1 \leq j \leq k+1$).
    Then,
    we consider $\vec{n_0} \in \Nat^k$ such that $\vec{n_0}$ has all 
    its components greater than $\factorial{s}$ and such that
    $w[\vec{n_0}]$ belongs to $L$. 
    Then, we inductively define 
    $\vec{n_{i+1}}$  as the smallest vector of numbers greater than $\vec{n_i}$,
    such that $w[\vec{n_{i+1}}]$ belongs to $L$, 
    and with $\vec{n_i}$ having all components greater than
    $2\card{w[\vec{n_i}]}$.

@@

    \todo[inline]{Aliaume: proof of concept proof}
    Let us now conclude that the same $x,y$ work for all $\vec{n} \in \Nat^k$.

@@

\begin{corollary}
    \label{bounded-wqo-dwclosed:cor}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the
    \kl{infix relation} if and only if $\dwset[\infleq]{L}$ is.
\end{corollary}
\begin{proof}
    Because $L \subseteq \dwset[\infleq]{L}$, the right-to-left implication
    is trivial.
    For the left-to-right implication, let us assume that $L$ is a
    \kl{well-quasi-ordered} language for the \kl{infix relation}.
    Then $L$ is included in a finite union 
    of products of \kl{chains} for the \kl{prefix} and \kl{suffix} relations
    thanks to
    \cref{bounded-language:thm}:
    \begin{equation*}
        L \subseteq \bigcup_{i = 1}^n S_i \cdot P_i \quad .
    \end{equation*}
    Remark that if $S_i$ is a \kl{chain} for the \kl{suffix relation}
    and $P_i$ is a \kl{chain} for the \kl{prefix relation},
    then 
    \begin{equation*}
      \dwset[\infleq]{(S_i \cdot P_i)} = (\dwset[\suffleq]{S_i}) \cdot (\dwset[\prefleq]{P_i})
        \quad .
    \end{equation*}
    Indeed, any \kl{infix} of a word in $S_i P_i$ can be split into
    a \kl{suffix} of a word in $S_i$ and a \kl{prefix} of a word in $P_i$.
    Conversely, any such concatenations are \kl{infixes} of a word in $S_i P_i$.

    As a consequence, we conclude that $\dwset[\infleq]{L}$ is itself included
    in a finite union of products of \kl{chains}.
    Furthermore, by definition of \kl{bounded languages},
    $\dwset[\infleq]{L}$ is also a \kl{bounded language}.
    Hence, it is \kl{well-quasi-ordered} by the \kl{infix relation} 
    via
    \cref{bounded-language:thm}.
\end{proof}

The \cref{bounded-wqo-dwclosed:cor} is reminiscent of a similar result for the
\kl{subword embedding},  stipulating that for any language $L \subseteq
\Sigma^*$, the \kl{downwards closure} $\dwset[\higleq]{L}$ is described using
finitely many excluded \kl{subwords}, hence is regular. However, this is not
the case for the \kl{infix relation}, even with \kl{bounded languages}, as we
will now illustrate with the following example.

\begin{example}
    \label{dwclosed-wqo-not-finite-excl:ex}
    Let $L \defined a^* b^* \cup b^* a^*$. This language is 
    \kl(language){bounded}, is \kl{downwards
    closed} for the \kl{infix relation}, is \kl{well-quasi-ordered} for the
    \kl{infix relation}, but is characterized by an \emph{infinite} number 
    of excluded infixes, respectively of the form $ab^ka$ and $ba^kb$ where $k \geq 1$.
\end{example}

@@

    Let us prove that these are upper bounds for the \kl{ordinal invariants} of
    $\LMorse$. The bound of the \kl{ordinal height} holds for any language $L$,
    as the length of a decreasing sequence of words is bounded by the length of
    its first element. For the \kl{maximal order type}, we remark that the
    \kl{uniform recurrence} of $\ThueMorse$ means that the maximal length of a
    \kl{bad sequence} is determined by its first element, hence that it is at
    most $\omegaOrd$. Finally, because the \kl{ordinal width} is at most the
    \kl{maximal order type} (as per \cref{ordinal-invariants:subsec}, using for instance the
    results of \cite{kriz90b} or \cite[Theorem 3.8]{DZSCSC20} stating 
    $\oType{X} \leq \oHeight{X} \oComProd \oWidth{X}$): we conclude
    that the \kl{ordinal width} is also at most $\omegaOrd$.

@@

\AP A subset $I \subseteq X$ is \intro(subset){directed} if, for every $x,y \in
I$, there exists $z \in I$ such that $x \leq z$ and $y \leq z$. Given a
\kl{well-quasi-order} $(X, \leq)$, one can always decompose $X$ into a finite
union of \intro{order ideals}, that is, non-empty sets $I \subseteq X$ that are
\kl{downwards closed} and \kl(subset){directed} for the relation $\leq$. In our
case, a \kl{well-quasi-ordered} \kl{order ideal} for the \kl{infix relation} is
the set of finite \kl{infixes} of a finite, infinite,
or bi-infinte word $w \in \Sigma^* \cup
\Sigma^\Nat \cup \Sigma^\Rel $ (\cref{bi-infinite:lem}). Therefore, we will
first study languages that are described as the set of finite \kl{infixes} of
some (bi-)infinite word (\cref{ultimately-uniformly-recurrent:lem}). To that
end, let us introduce the notation $\intro*\infset{w}$ for the set of finite
\kl{infixes} of a (possibly infinite or bi-infinite) word $w$.

### 2025-10-13 wqo-on-words.lncs.tex
fix: correct typos and error noticed by Lia

% math packages
\usepackage{amssymb,amsmath,stmaryrd,thmtools,thm-restate,upgreek}

@@

% links inside the document
\usepackage{hyperref}
\usepackage[capitalise,noabbrev,nameinlink]{cleveref}
\Crefname{assumption}{Assumption}{Assumptions}
\Crefname{conjecture}{Conjecture}{Conjectures}
\usepackage[composition,hyperref,xcolor,cleveref]{knowledge}
\knowledgeconfigure{notion}

@@

\author{
        Nathan Lhote\inst{3}\orcidID{0000-0003-3303-5368} \and
        Aliaume Lopez\inst{2}\orcidID{0000-0002-4205-327X} \and
        Lia Schütze\inst{1}\orcidID{0000-0003-4002-5491}}
\authorrunning{N. Lhote and A. Lopez and L. Schütze}
\institute{
Max Planck Institute for Software Systems
 \and
INP Bordeaux, LaBRI, CNRS
 \and
Aix-Marseille University
}

>>>>

% math packages
\usepackage{amssymb,amsmath,amsthm,stmaryrd,thmtools,upgreek}

@@

% links inside the document
\usepackage{hyperref}
\usepackage[capitalise,noabbrev,nameinlink]{cleveref}
\Crefname{assumption}{Assumption}{Assumptions}
\usepackage[composition,hyperref,xcolor,cleveref]{knowledge}
\knowledgeconfigure{notion}

@@

\author{
        Nathan Lhote\inst{2}\orcidID{0000-0003-3303-5368} \and
        Aliaume Lopez\inst{3}\orcidID{0000-0002-4205-327X} \and
        Lia Schütze\inst{1}\orcidID{0000-0003-4002-5491}}
\authorrunning{N. Lhote and A. Lopez and L. Schütze}
\institute{
Max Planck Institute for Software Systems
 \and
Aix-Marseille University
 \and
INP Bordeaux, LaBRI, CNRS
}
